\textbf{Towards transparency of hybrid open access through publisher-provided metadata: An article-level study of Elsevier}

### 2021-05-19 analysis/manuscript.Rmd
align text

We identified OA articles
through CC licenses in Crossref metadata records (Hendricks et al., 2020) and then downloaded the XML
version of all CC-licensed articles published in a hybrid journal. From
the XML files, we obtained the articles' OA status and invoicing information (see Table \ref{tab:els_xml}). We determined whether articles were immediate or delayed
OA using the XML node \texttt{openArchiveArticle}, which indicates if an article was made freely available under Elsevier's Open Archive program, and measured the
uptake of hybrid OA. Moreover, we obtained the invoice channels and recipients of hybrid OA APCs. Using the \texttt{openaccessSponsorType} node, we distinguished between four invoice channels, including
invoices billed to authors, as part of publishing
agreements with funding bodies (cf. @Els_Agreements), exempted through fee waivers (e.g., in ``cases of genuine need'' or due to
society or university sponsorships, cf.~Elsevier,
n.d.-c), and other types not
specified by Elsevier. Finally, when hybrid APCs were invoiced as part
of agreements, we identified invoice recipients through the
\texttt{openaccessSponsorName} node. It is worth noting that while this article focuses on OA articles published in hybrid journals, the same data is also available for articles published in full OA journals. Elsevier does not provide the invoiced APC in either case.

@@

Figure \ref{fig:boxuptake} presents the hybrid OA uptake by year, showing considerable variation among Elsevier's hybrid journals. The box-and-whiskers plot visualizes the median hybrid OA uptake (horizontal line inside the box), the range of the majority of journals (the box), and individual outliers (dots). Notably, the upper quartiles and whiskers stretch farther over the years, which indicates that the range of uptake among journals with an above-average proportion of OA articles increased over the years. Nevertheless, the prevalence of hybrid OA remained relatively low; in 2019, 95% of Elsevier hybrid journals published ≤`r perc_df %>% filter(issued_year == 2019) %>% pull(p_95) * 100`% of their articles OA.

```{r boxuptake, fig.cap = "Open access uptake per Elsevier hybrid journal in percent by year, visualized as box-and-whiskers plot. The Y-axis is limited to an open access share of 20\\%."}
hybrid_oa_volume %>%
  mutate(prop = oa / articles) %>%
  filter(oa > 0) %>%
ggplot(aes(factor(issued_year), prop)) + 
  geom_boxplot(outlier.size = 1, outlier.alpha = .5) + 
  coord_cartesian(ylim=c(0,0.2)) +
  theme_minimal_hgrid() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 5L),
      expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Publication Year",
       y = "Hybrid OA Uptake")
```

### Share of the Total Volume published in OA{.unnumbered}

@@

During the same period, Elsevier published `r all_oa` OA articles in total (i.e., in hybrid and full OA journals), representing `r all_oa / all_els * 100`% of the total article volume (n=`r all_els`). Looking at Figure \ref{fig:oa_volume_fig}, it is apparent that the OA article volume of hybrid journals (n=`r filter(summary_oa_type, type == "hybrid_immediate_articles") %>% pull(n)`; `r filter(summary_oa_type, type == "hybrid_immediate_articles") %>% pull(prop) * 100`%) lagged behind that of Elsevier's Open Archive program (n=`r filter(summary_oa_type, type == "hybrid_delayed_articles") %>% pull(n)`; `r filter(summary_oa_type, type == "hybrid_delayed_articles") %>% pull(prop) * 100`%) and full OA journals (n=`r filter(summary_oa_type, type == "full_oa") %>% pull(n)`; `r filter(summary_oa_type, type == "full_oa") %>% pull(prop) * 100`%), which included `r sum(mirror_jn$all_articles)` articles published in `r length(unique(mirror_jn$container_title))` mirror journals.

```{r oa_volume_fig, fig.cap="Open Access (OA) in Elsevier journals in terms of article volume between 2015-2019. Figure A presents the aggregated OA article volume of Elsevier journals from 2015-2019 by OA type. Figure B presents the share of OA articles relative to Elsevier's overall article output by OA type and open content license from 2015-2019."}
# see analysis/006_license.Rmd how the figure was created

@@

As far as we observed, OA articles in Elsevier hybrid journals were published under two possible CC licenses--CC BY or CC BY-NC-ND. CC BY allows others to distribute, remix, adapt, and build upon the licensed work, including for commercial purposes, as long as the original author is credited. In contrast, CC BY-NC-ND is less permissive as it prohibits commercial reuse and derivatives. In the five-year period from 2015-2019, the proportion of hybrid OA articles published under a CC BY-NC-ND license marginally increased and maintained a greater share than that of CC BY (see Figure \ref{fig:oa_volume_fig}-B). Interestingly, Figure \ref{fig:oa_volume_fig}-B also reveals that hybrid journals had the highest number and proportion of openly available articles licensed under CC BY (n=`r filter(cc_by_by_type, type == "hybrid_immediate_articles") %>% pull(n)`; `r filter(cc_by_by_type, type == "hybrid_immediate_articles") %>% pull(prop) *100`%) compared to full OA journals  (n=`r filter(cc_by_by_type, type == "full_oa") %>% pull(n)`; `r filter(cc_by_by_type, type == "full_oa") %>% pull(prop) *100`%) and Elsevier's delayed OA program (n=`r filter(cc_by_by_type, type == "hybrid_delayed_articles") %>% pull(n)`; `r filter(cc_by_by_type, type == "hybrid_delayed_articles") %>% pull(prop) *100`%) . Most  delayed OA articles were provided under an Elsevier user license (Els-User), which prohibits reuse for commercial purposes.

@@

Next, we present disciplinary variations in hybrid OA. Table \ref{tab:subject_area_table} presents the high-level findings by AJSC subject area. In the five-year period we studied, the physical sciences (`r filter(subject_area_all_jn, subject_area == "Physical Sciences") %>% pull(journals_n)` hybrid journals, `r filter(subject_area_all_article, subject_area == "Physical Sciences") %>% pull(articles_n)` OA articles) and health sciences (`r filter(subject_area_all_jn, subject_area == "Health Sciences") %>% pull(journals_n)` hybrid journals, `r filter(subject_area_all_article, subject_area == "Health Sciences") %>% pull(articles_n)` OA articles) were the most prevalent in terms of hybrid journals with at least one OA article.  In terms of article volume, most OA articles were published in hybrid journals belonging to the life sciences (`r filter(subject_area_all_jn, subject_area == "Life Sciences") %>% pull(journals_n)` hybrid journals, `r filter(subject_area_all_article, subject_area == "Life Sciences") %>% pull(articles_n)` OA articles), while social sciences  (`r filter(subject_area_all_jn, subject_area == "Social Sciences") %>% pull(journals_n)`  hybrid journals, `r filter(subject_area_all_article, subject_area == "Social Sciences") %>% pull(articles_n)` articles) journals published the lowest number of OA articles. However, it is important to note that there is a large overlap between the journals’ subject areas. Therefore, Table  \ref{tab:subject_area_table} also assigns hybrid journals and OA articles fractionally to a subject area and compares it to the full counting, where journals assigned more than one subject in the AJSC classification were counted once for each field. For the remaining results, we only present the full counting.

```{r subject_area_table}
# join
inner_join(subject_area_all_jn, subject_area_all_article, by = "subject_area") %>%
  arrange(subject_area) %>%
  select(1, journals_n, articles_n, journals_frac, articles_frac) %>%
  filter(subject_area != "Multidisciplinary", !is.na(subject_area)) %>% 
  mutate(subject_area = stringr::str_to_sentence(subject_area)) %>%
  kbl(booktabs = TRUE, col.names = c("Subject area", "Journals", "OA articles", "Journals", "OA articles"),
      caption = "Subject area overview using full and fractional counting. Table reports the number of Elsevier hybrid journals with at least one open access (OA) article, and OA articles published from 2015-2019. One multidisciplinary with 61 OA articles and 11 journals with 37 OA articles that  could not be matched to the Scopus Source Title list used to obtain ASJC codes are not included.",
      format.args = list(big.mark = ",")) %>%
  add_header_above(c(" " = 1, "Full counting" = 2, "Fractional counting" = 2)) %>%
  kable_styling() 
```

@@

Figure \ref{fig:oa_sub_uptake} presents the journals' five-year OA uptake grouped by subject area and field and in comparison to the overall median (Mdn= `r median_all * 100`; excluding one journal coded as multidisciplinary). The box plot reveals some variation between the subject areas. Notably, with the exception of environmental science and earth and planetary sciences, journals from the physical sciences show a lower median proportion of hybrid OA articles, whereas most life sciences and social sciences journals ranked above average. Figure X also shows large variations among the 26 subject fields, with the highest OA rates in immunology and microbiology (Mdn=`r filter(oa_per_subject_rank, description == "Immunology and Microbiology") %>% pull(median_prop)*100`), environmental science (Mdn=`r filter(oa_per_subject_rank, description == "Environmental Science") %>% pull(median_prop)*100`), and neuroscience (Mdn=`r filter(oa_per_subject_rank, description == "Neuroscience") %>% pull(median_prop)*100`), whilst  chemistry (Mdn=`r filter(oa_per_subject_rank, description == "Chemistry") %>% pull(median_prop)*100`), materials science (Mdn=`r filter(oa_per_subject_rank, description == "Materials Science") %>% pull(median_prop)*100`), and dentistry (Mdn=`r filter(oa_per_subject_rank, description == "Dentistry") %>% pull(median_prop)*100`) recorded the lowest OA rate.

```{r oa_sub_uptake, fig.cap='Open access (OA) uptake in Elsevier hybrid journals in percent grouped by subject area and field between 2015-19. Visualized as a box-and-whiskers plot limited to an OA share of 20\\%; the red line represents the overall median (2.5), jittered dots represent individual journals. Subject fields are ordered based on the median OA uptake.', fig.asp=1.3, out.width="90%"}
# with labels
subject_all_years_ <- 
  inner_join(subject_all_years, journals_per_subject, by = "description")

@@

### Citation Impact{.unnumbered}

```{r}
hybrid_subject_19 <- hybrid_subject %>% 
  mutate(prop = oa /articles) %>%
  filter(issued_year == "2019", !is.na(percentile)
         , prop != 0) %>%
  distinct(issn, prop, percentile, snip, sjr)

cor_test_res <- cor.test(hybrid_subject_19$prop, hybrid_subject_19$percentile, method = "spearman") 
```

Furthermore, we were interested in the relationship between the journals' field-specific impact and their OA uptake. Spearman's rho correlation coefficient was used to assess the relationship between the 2019 Source Normalized Impact per Paper (SNIP) value calculated by Scopus and the journals' OA uptake that year. Considering only journals with at least one OA article, we found a weak but positive correlation between journal impact and OA uptake ($r_s$=`r as.character(round(cor_test_res$estimate,4))`, $p<0.001$), suggesting that the relationship is not very strong, based on our data (see Figure \ref{fig:sniptest}).

```{r sniptest, fig.cap="Source Normalized Impact per Paper (SNIP) versus open access (OA) uptake 2019. Each dot represents a hybrid journal with at least one published OA article in 2019. There is only a weak, positive correlation between the two variables ($r_s$=0.1679, $p<0.001$)"}
ggplot(hybrid_subject_19, aes(prop, snip)) + geom_point(alpha = 0.2) + 
  geom_smooth(method = "loess", se = TRUE) +
  coord_cartesian(xlim = c(0., 0.7), ylim = c(0.5, 6)) +
  labs(x = "Hybrid OA Uptake", y = "SNIP") +
  theme_cowplot()
```

@@

As can be seen from Table \ref{tab:invoicing_overview}, hybrid OA APC were most often invoiced to authors (n=`r filter(sponsored_art, fct_source == "Author") %>% pull(nn)`; `r filter(sponsored_art, fct_source == "Author") %>% pull(prop)`%) and to a lesser extent as part of agreements (n=`r filter(sponsored_art, fct_source == "Agreement") %>% pull(nn)`; `r filter(sponsored_art, fct_source == "Agreement") %>% pull(prop)`%). Interestingly, we also found a small number of cases where hybrid APCs were waived (n=`r filter(sponsored_art, fct_source == "Fee waived") %>% pull(nn)`; `r filter(sponsored_art, fct_source == "Fee waived") %>% pull(prop)`%).

```{r invoicing_overview}
sponsored_art %>% 
  kbl(booktabs = TRUE, col.names = c("Invoicing channel", "Hybrid OA articles (n)", "Percentage"),
      caption = "Invoice channels for hybrid OA in Elsevier journals between 2015-19.", digits = 1,
      format.args = list(big.mark = ",")) %>%
  kable_styling() %>%
  row_spec(4, hline_after = TRUE)
```

@@

Figure \ref{fig:invoiceoverview} illustrates that over the years, Elsevier has constantly invoiced hybrid OA APCs directly to authors compared to research funders or academic consortia ("Agreement"). Between 2015 and 2019, the proportion of author-invoicing increased from `r filter(sponsored_articles_prop, issued_year == 2015, fct_source == "Author") %>% pull(perc)`% to `r filter(sponsored_articles_prop, issued_year == 2019, fct_source == "Author") %>% pull(perc)`%, while the share of invoicing through agreements slightly changed from `r filter(sponsored_articles_prop, issued_year == 2015, fct_source == "Agreement") %>% pull(perc)`% to `r filter(sponsored_articles_prop, issued_year == 2019, fct_source == "Agreement") %>% pull(perc)`%. Also, the share of fee-waived articles remained relatively stable, but we found different types of waivers. Around `r filter(els_waived_df, !is.na(oa_sponsor_name)) %>% distinct(doi) %>% nrow() / nrow(els_waived_df)* 100`% of fee waivers were linked to an OA sponsor, which indicates that a third party paid the APC. For instance, `r filter(els_waived_df, oa_sponsor_name == "Académie des Sciences") %>% distinct(doi) %>% nrow()` waivers were linked to the French Académie des Sciences, presumably covering OA publication in its society journals for affiliated authors. The remaining `r filter(els_waived_df, is.na(oa_sponsor_name)) %>% distinct(doi) %>% nrow() / nrow(els_waived_df)*100`% of waived articles did not disclose any OA sponsor. Moreover, Figure \ref{fig:invoiceoverview} compares the invoicing channels based on open content license variants. When Elsevier invoiced authors directly, most OA articles in hybrid journals were published under a non-commercial license (n=`r filter(sponsored_articles, fct_source == "Author", cc_norm == "CC BY-NC-ND") %>% pull(n) %>% sum()`; `r filter(sponsored_articles, fct_source == "Author", cc_norm == "CC BY-NC-ND") %>% pull(n) %>% sum() / filter(sponsored_articles, fct_source == "Author") %>% pull(n) %>% sum()* 100`%), whereas most articles billed as part of agreements were licensed under the more permissive CC BY license (n=`r filter(sponsored_articles, fct_source == "Agreement", cc_norm == "CC BY") %>% pull(n) %>% sum()`; `r filter(sponsored_articles, fct_source == "Agreement", cc_norm == "CC BY") %>% pull(n) %>% sum() / filter(sponsored_articles, fct_source == "Agreement") %>% pull(n) %>% sum() * 100`%).

```{r invoiceoverview, fig.cap="Development of fee-based open access (OA) publishing in Elsevier hybrid journals by invoice channel Colored bars represent the share of OA articles invoiced to authors, agreements, or waived, grouped by Creative Commons license variant. Grey bars show the total number of hybrid OA articles published in Elsevier journals from 2015-2019."}
# plot
sponsored_articles %>%
  mutate(fct_source = stringr::str_to_title(fct_source)) %>%
  mutate(fct_source = ordered(fct_source, levels = c("Author", "Agreement", "Fee Waived", "Other"))) %>%
  ggplot(aes(x = gsub("20", "'", issued_year), y = n), alpha = 0.8) +
  geom_bar(
    data = all_articles,
    color = "transparent",
    stat = "identity",
    fill = "#b3b3b3a0"
  ) +
  geom_bar(aes(fill = cc_norm), color = "white", stat = "identity", alpha = 0.8, size = 0.25,
           position = position_stack(reverse = T)) +
  facet_wrap( ~ fct_source, nrow = 1) +
  scale_fill_manual(values = 
                      c(`CC BY` = "#B52141",
                        `CC BY-NC-ND` = "#0093c7"),
                    name = "CC License Variant") +
  labs(x = "Publication Year", y = "Hybrid OA Articles") + 
  scale_y_continuous( 
                     labels = scales::number_format(big.mark = ","),
                      expand = expansion(mult = c(0, 0.05))) +
  theme_minimal_hgrid() +
  theme(legend.position="top", legend.justification = "right")
```

We also observed large differences in OA invoicing among subject fields. Table \ref{tab:invocing_by_subject} shows the number of OA articles by subject field and invoice channel. For articles in nursing, decision dciences, and pharmacology, toxicology and pharmaceutics, Elsevier predominantly invoiced authors, whereas most energy and chemical engineering articles were invoiced through agreements. Likewise, the majority of articles in materials science, chemistry and physics and astronomy were not invoiced to authors but facilitated through agreements or waived. The large share of waived APCs in physics and astronomy can be attributed to a single 2015 issue of Nuclear and Particle Physics Proceedings, which offered sponsored OA.

@@

Elsevier’s data offer more insight into invoicing. Overall, we identified 62 sponsoring institutions and organizations that received invoices for hybrid publishing fees as part of an agreement with Elsevier. We found that, by a large margin, most invoices were issued to UK-based research funders and institutions (n=`r filter(sponsor_df, country_factor == "United Kingdom") %>% distinct(doi) %>% nrow()`; `r filter(sponsor_df, country_factor == "United Kingdom") %>% distinct(doi) %>% nrow() / sponsor_df %>% distinct(doi) %>% nrow()* 100`%), followed by the Netherlands (n=`r filter(sponsor_df, country_factor == "The Netherlands") %>% distinct(doi) %>% nrow()`; `r filter(sponsor_df, country_factor == "The Netherlands") %>% distinct(doi) %>% nrow() / sponsor_df %>% distinct(doi) %>% nrow()* 100`%) and the European Union (n=`r filter(sponsor_df, country_factor == "European Union") %>% distinct(doi) %>% nrow()`; `r filter(sponsor_df, country_factor == "European Union") %>% distinct(doi) %>% nrow() / sponsor_df %>% distinct(doi) %>% nrow()* 100`%; EU). Figure \ref{fig:invoice_sponsor_country} provides a closer look at the geographical distribution by license prevalence, highlighting the dominance of CC -BY licensed articles invoiced to sponsoring institutions in the UK, the United States, and, to a much lesser extent, the Netherlands. In contrast, the majority of articles invoiced to sponsoring institutions from Norway or representing the EU were published under a non-commercial license.

>>>>

We identified OA articles
through CC licenses in Crossref metadata records (Hendricks et al., 2020) and then downloaded the XML
version of all CC-licensed articles published in a hybrid journal. From
the XML files, we obtained the articles' OA status and invoicing information (see Table \ref{tab:els_xml}). We determined whether articles were immediate or delayed
OA using the XML node \texttt{openArchiveArticle}, which indicates if an article was made freely available under Elsevier's Open Archive program, and measured the
uptake of hybrid OA. Moreover, we obtained the invoice channels and recipients of hybrid OA APCs. Using the \texttt{openaccessSponsorType} node, we distinguished between four invoice channels, including
invoices billed to authors, as part of publishing
agreements with funding bodies (cf. @Els_Agreements), exempted through fee waivers (e.g., in ``cases of genuine need'' or due to
society or university sponsorships, cf. @Els_Pricing), and other types not
specified by Elsevier. Finally, when hybrid APCs were invoiced as part
of agreements, we identified invoice recipients through the
\texttt{openaccessSponsorName} node. It is worth noting that while this article focuses on OA articles published in hybrid journals, the same data is also available for articles published in full OA journals. Elsevier does not provide the invoiced APC in either case.

@@

Table \ref{tab:overview} also shows that the average OA share per journal was higher than the overall percentage. To consider these variations across journals, Figure \ref{fig:boxuptake} illustrates the range of OA uptake rates as a box-and-whiskers plot by year. The bold line shows the median OA prevalence across journals, which slightly rose from `r filter(perc_df, issued_year == 2015) %>% pull(median) * 100`% in 2015 to `r filter(perc_df, issued_year == 2019) %>% pull(median) * 100`% in 2019}. Notably, the upper quartiles and
whiskers stretch farther over the years, which indicates that the range
of uptake among journals with an above-average proportion of OA articles
increased over the years. Nevertheless, the prevalence of hybrid OA
remained relatively low; in 2019, 95% of Elsevier hybrid journals published ≤`r perc_df %>% filter(issued_year == 2019) %>% pull(p_95) * 100`% of their articles OA.

```{r boxuptake, fig.cap = "Open Access Uptake per Elsevier Hybrid Journal (Per Year)."}
hybrid_oa_volume %>%
  mutate(prop = oa / articles) %>%
  filter(oa > 0) %>%
ggplot(aes(factor(issued_year), prop)) + 
  geom_boxplot(outlier.size = 1, outlier.alpha = .5) + 
  coord_cartesian(ylim=c(0,0.2)) +
  theme_minimal_hgrid() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 5L),
      expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Publication Year",
       y = "Hybrid OA Uptake")
```

@@

During the same period, Elsevier published OA to `r all_oa` articles in total (i.e., full, hybrid, and delayed OA combined), representing `r all_oa / all_els * 100`% of the total article volume (n=`r all_els`). Looking at Figure \ref{fig:oa_volume_fig}, it is apparent that the OA article volume of hybrid journals (n=`r filter(summary_oa_type, type == "hybrid_immediate_articles") %>% pull(n)`; `r filter(summary_oa_type, type == "hybrid_immediate_articles") %>% pull(prop) * 100`%) lagged delayed OA (n=`r filter(summary_oa_type, type == "hybrid_delayed_articles") %>% pull(n)`; `r filter(summary_oa_type, type == "hybrid_delayed_articles") %>% pull(prop) * 100`%) and full OA journals (n=`r filter(summary_oa_type, type == "full_oa") %>% pull(n)`; `r filter(summary_oa_type, type == "full_oa") %>% pull(prop) * 100`%), which included `r sum(mirror_jn$all_articles)` articles published in `r length(unique(mirror_jn$container_title))` mirror journals.

```{r oa_volume_fig, fig.cap="Elsevier OA Article Volume 2015-2019 by (A) OA Type and (B) Open Content License."}
# see analysis/006_license.Rmd how the figure was created

@@

As far as we observed, OA articles in Elsevier hybrid journals were published under two possible CC licenses--CC BY or CC BY-NC-ND. CC BY allows others to distribute, remix, adapt, and build upon the licensed work, including for commercial purposes, as long as the original author is credited. In contrast, CC BY-NC-ND is less permissive as it prohibits commercial reuse and derivatives. In the five-year period from 2015-2019, the proportion of hybrid OA articles published under a CC BY-NC-ND license marginally increased and maintained a greater share than that of CC BY (see Figure \ref{fig:oa_volume_fig}-B). Interestingly, Figure \ref{fig:oa_volume_fig}-B also reveals that hybrid journals had the highest number and proportion of openly available articles licensed under CC BY (n=`r filter(cc_by_by_type, type == "hybrid_immediate_articles") %>% pull(n)`; `r filter(cc_by_by_type, type == "hybrid_immediate_articles") %>% pull(prop) *100`%) compared to full OA journals  (n=`r filter(cc_by_by_type, type == "full_oa") %>% pull(n)`; `r filter(cc_by_by_type, type == "full_oa") %>% pull(prop) *100`%) and Elsevier's delayed OA program (n=`r filter(cc_by_by_type, type == "hybrid_delayed_articles") %>% pull(n)`; `r filter(cc_by_by_type, type == "hybrid_delayed_articles") %>% pull(prop) *100`%) . Most  delayed OA articles were provided under an Elsevier user license (Els-User), which prohibits reuse for commercial purposes and redistribution.

@@

Next, we present the hybrid OA uptake of different disciplines. Table
\ref{tab:subject_area_table} presents the high-level findings by ASJC
subject area. Between 2015 and 2019, the physical sciences
(`r filter(subject_area_all_jn, subject_area == "Physical Sciences") %>% pull(journals_n)` hybrid journals, `r filter(subject_area_all_article, subject_area == "Physical Sciences") %>% pull(articles_n)` OA articles) and health sciences  (`r filter(subject_area_all_jn, subject_area == "Health Sciences") %>% pull(journals_n)` hybrid journals, `r filter(subject_area_all_article, subject_area == "Health Sciences") %>% pull(articles_n)` OA articles) recorded the most
hybrid journals with at least one OA article. However,
the life sciences ((`r filter(subject_area_all_jn, subject_area == "Life Sciences") %>% pull(journals_n)` hybrid journals, `r filter(subject_area_all_article, subject_area == "Life Sciences") %>% pull(articles_n)` OA articles) published the most hybrid OA articles, while the
social sciences published the least (`r filter(subject_area_all_jn, subject_area == "Social Sciences") %>% pull(journals_n)`  hybrid journals, `r filter(subject_area_all_article, subject_area == "Social Sciences") %>% pull(articles_n)` articles). It is important to
note, though, that there is a large overlap between the journals' subject areas. Therefore, Table \ref{tab:subject_area_table} also presents fractional counts for hybrid journals and OA articles, which equally weights counting with respect to the total number of ASJC subject areas a journal belongs to. Journals with multiple ASJC codes that are members of the same subject area were counted once. For the remaining results, we only present the full counting.

```{r subject_area_table}
# join
inner_join(subject_area_all_jn, subject_area_all_article, by = "subject_area") %>%
  arrange(subject_area) %>%
  select(1, journals_n, articles_n, journals_frac, articles_frac) %>%
  filter(subject_area != "Multidisciplinary", !is.na(subject_area)) %>% 
  mutate(subject_area = stringr::str_to_sentence(subject_area)) %>%
  kbl(booktabs = TRUE, col.names = c("Subject area", "Journals", "OA articles", "Journals", "OA articles"),
      caption = "Full and Fractional Counting of Subject Areas (2015-2019).",
      format.args = list(big.mark = ",")) %>%
  add_header_above(c(" " = 1, "Full counting" = 2, "Fractional counting" = 2)) %>%
  kable_styling() 
```

@@

Figure \ref{fig:oa_sub_uptake} presents the journals' five-year OA uptake grouped by subject area and field and in comparison to the overall median (Mdn= `r median_all * 100`; excluding one journal coded as multidisciplinary). The box plot reveals some variation between the subject areas. Notably, with the exception of environmental science and earth and planetary sciences, journals from the physical sciences show a lower median proportion of hybrid OA articles, whereas most life sciences and social sciences journals ranked above average. Figure \ref{fig:oa_sub_uptake}  also shows large variations among the 26 subject fields, with the highest OA rates in immunology and microbiology (Mdn=`r filter(oa_per_subject_rank, description == "Immunology and Microbiology") %>% pull(median_prop)*100`), environmental science (Mdn=`r filter(oa_per_subject_rank, description == "Environmental Science") %>% pull(median_prop)*100`), and neuroscience (Mdn=`r filter(oa_per_subject_rank, description == "Neuroscience") %>% pull(median_prop)*100`), whilst  chemistry (Mdn=`r filter(oa_per_subject_rank, description == "Chemistry") %>% pull(median_prop)*100`), materials science (Mdn=`r filter(oa_per_subject_rank, description == "Materials Science") %>% pull(median_prop)*100`), and dentistry (Mdn=`r filter(oa_per_subject_rank, description == "Dentistry") %>% pull(median_prop)*100`) recorded the lowest OA rate.

```{r oa_sub_uptake, fig.cap='OA Uptake in Elsevier Hybrid Journals by Subject Area and Field (2015-2019).', fig.asp=1.3, out.width="90%"}
# with labels
subject_all_years_ <- 
  inner_join(subject_all_years, journals_per_subject, by = "description")

@@

@@

As can be seen from Table \ref{tab:invoicing_overview}, hybrid OA APCs were most often invoiced to authors (n=`r filter(sponsored_art, fct_source == "Author") %>% pull(nn)`; `r filter(sponsored_art, fct_source == "Author") %>% pull(prop)`%) and to a lesser extent as part of agreements (n=`r filter(sponsored_art, fct_source == "Agreement") %>% pull(nn)`; `r filter(sponsored_art, fct_source == "Agreement") %>% pull(prop)`%). Interestingly, we also found a small number of cases where hybrid APCs were waived (n=`r filter(sponsored_art, fct_source == "Fee waived") %>% pull(nn)`; `r filter(sponsored_art, fct_source == "Fee waived") %>% pull(prop)`%). A brief note of caution regarding APCs invoiced to the authors is due here: Although author invoicing was indicated in the \texttt{openaccessSponsorType} article metadata field used for this study, this does not necessarily mean that authors paid the APC out of their own pocket. Rather, this implies that the authors' institution or funder did not have a central invoicing agreement in place at the time of publication.

```{r invoicing_overview}
sponsored_art %>% 
  kbl(booktabs = TRUE, col.names = c("Invoicing channel", "Hybrid OA articles (n)", "Percentage"),
      caption = "Invoice Channels for Elsevier Hybrid OA Articles (2015-2019).", digits = 1,
      format.args = list(big.mark = ",")) %>%
  kable_styling() %>%
  row_spec(4, hline_after = TRUE)
```

@@

Figure \ref{fig:invoiceoverview} illustrates that over the years, Elsevier has consistently invoiced more authors than funders or academic consortia ("Agreement"), although both shares increased. In particular, the proportion of author-invoicing increased from `r filter(sponsored_articles_prop, issued_year == 2015, fct_source == "Author") %>% pull(perc)`% to `r filter(sponsored_articles_prop, issued_year == 2019, fct_source == "Author") %>% pull(perc)`%, whereas the share of invoicing through agreements slightly
increased from `r filter(sponsored_articles_prop, issued_year == 2015, fct_source == "Agreement") %>% pull(perc)`% to `r filter(sponsored_articles_prop, issued_year == 2019, fct_source == "Agreement") %>% pull(perc)`%, respectively. The share of fee-waived articles also remained relatively stable, but we found different types of waivers. Around `r filter(els_waived_df, !is.na(oa_sponsor_name)) %>% distinct(doi) %>% nrow() / nrow(els_waived_df)* 100`% of fee waivers were linked to an OA sponsor, which indicates that a third party paid the APC. For instance, `r filter(els_waived_df, oa_sponsor_name == "Académie des Sciences") %>% distinct(doi) %>% nrow()` waivers were linked to the French Académie des Sciences, presumably covering OA publication in its society journals for affiliated authors. The remaining `r filter(els_waived_df, is.na(oa_sponsor_name)) %>% distinct(doi) %>% nrow() / nrow(els_waived_df)*100`% of waived articles did not disclose any OA sponsor. Moreover, Figure \ref{fig:invoiceoverview} compares the invoicing channels based on open content license variants. When Elsevier invoiced authors directly, most OA articles in hybrid journals were published under a non-commercial license (n=`r filter(sponsored_articles, fct_source == "Author", cc_norm == "CC BY-NC-ND") %>% pull(n) %>% sum()`; `r filter(sponsored_articles, fct_source == "Author", cc_norm == "CC BY-NC-ND") %>% pull(n) %>% sum() / filter(sponsored_articles, fct_source == "Author") %>% pull(n) %>% sum()* 100`%), whereas most articles billed as part of agreements were licensed under the more permissive CC BY license (n=`r filter(sponsored_articles, fct_source == "Agreement", cc_norm == "CC BY") %>% pull(n) %>% sum()`; `r filter(sponsored_articles, fct_source == "Agreement", cc_norm == "CC BY") %>% pull(n) %>% sum() / filter(sponsored_articles, fct_source == "Agreement") %>% pull(n) %>% sum() * 100`%).

```{r invoiceoverview, fig.cap="Elsevier Hybrid OA Articles by Invoice Channel and CC License (Per Year)."}
# plot
sponsored_articles %>%
  mutate(fct_source = stringr::str_to_title(fct_source)) %>%
  mutate(fct_source = ordered(fct_source, levels = c("Author", "Agreement", "Fee Waived", "Other"))) %>%
  ggplot(aes(x = gsub("20", "'", issued_year), y = n), alpha = 0.8) +
  geom_bar(
    data = all_articles,
    color = "transparent",
    stat = "identity",
    fill = "#b3b3b3a0"
  ) +
  geom_bar(aes(fill = cc_norm), color = "white", stat = "identity", alpha = 0.8, size = 0.25,
           position = position_stack(reverse = T)) +
  facet_wrap( ~ fct_source, nrow = 1) +
  scale_fill_manual(values = 
                      c(`CC BY` = "#B52141",
                        `CC BY-NC-ND` = "#0093c7"),
                    name = "CC License Variant") +
  labs(x = "Publication Year", y = "Hybrid OA Articles") + 
  scale_y_continuous( 
                     labels = scales::number_format(big.mark = ","),
                      expand = expansion(mult = c(0, 0.05))) +
  theme_minimal_hgrid() +
  theme(legend.position="top", legend.justification = "right")
```

We also observed large differences in OA invoicing among subject fields.
Table \ref{tab:invocing_by_subject} shows the number of OA articles by
subject field and invoice channel. For articles in nursing, decision
sciences, and pharmacology, toxicology and pharmaceutics, Elsevier
predominantly invoiced authors, whereas most energy and chemical
engineering articles were invoiced through agreements. Likewise, the
majority of articles in materials science, chemistry and physics and
astronomy were not invoiced to authors but facilitated through
agreements or waived. The large share of waived APCs in physics and
astronomy can be attributed to a single 2015 issue of Nuclear and
Particle Physics Proceedings.

@@

Elsevier’s data offer more insight into invoicing. Overall, we identified `r filter(sponsor_df, !is.na(oa_sponsor_cleaned)) %>% distinct(oa_sponsor_cleaned) %>% nrow()` academic institutions and funders that received
invoices as part of publishing agreements. By a large margin, most invoices were issued to
UK-based research funders and institutions (n=`r filter(sponsor_df, country_factor == "United Kingdom") %>% distinct(doi) %>% nrow()`; `r filter(sponsor_df, country_factor == "United Kingdom") %>% distinct(doi) %>% nrow() / sponsor_df %>% distinct(doi) %>% nrow()* 100`%), followed by the Netherlands (n=`r filter(sponsor_df, country_factor == "The Netherlands") %>% distinct(doi) %>% nrow()`; `r filter(sponsor_df, country_factor == "The Netherlands") %>% distinct(doi) %>% nrow() / sponsor_df %>% distinct(doi) %>% nrow()* 100`%) and the European Union (n=`r filter(sponsor_df, country_factor == "European Union") %>% distinct(doi) %>% nrow()`; `r filter(sponsor_df, country_factor == "European Union") %>% distinct(doi) %>% nrow() / sponsor_df %>% distinct(doi) %>% nrow()* 100`%; EU), which in the context of this paper refers to institutions established by the European Commission, and not the  individual member states. Invoicing data comprises two EU institutions, the research funder European Research Council and the health agency European Center for Disease Prevention and Control (ECDS). Figure \ref{fig:invoice_sponsor_country} presents the geographical distribution by license prevalence,
highlighting the dominance of CC BY licensed articles invoiced to institutions in the UK, the United States (US), and, to a much
lesser extent, the Netherlands. In contrast, most articles
invoiced to institutions from Norway or representing the EU
were published under a non-commercial license.

```{r invoice_sponsor_country, fig.cap="Centrally Invoiced Elsevier Hybrid OA Articles by Country and CC License (2015-2019)."}
sponsor_df %>%
  group_by(country_factor, cc_norm) %>% 
  summarise(n = n_distinct(doi)) %>%
  ggplot(aes(fct_rev(country_factor), n, fill = cc_norm)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE),
            color = "white", alpha = 0.8, size = 0.25) +
  coord_flip() +
  scale_fill_manual(values = 
                      c(`CC BY` = "#B52141",
                        `CC BY-NC-ND` = "#0093c7"),
                    name = "CC License Variant") +
  labs(x = "OA Sponsor by Country", y = "Hybrid OA Articles (Agreement)") +
  scale_y_continuous(labels = scales::number_format(big.mark = ","),
                      expand = expansion(mult = c(0, 0.05)),
                     breaks = seq(0,15000, 2000)) +
  theme_minimal_vgrid() +
  theme(legend.position="top", legend.justification = "right")
```
